# Reproducible Research: Peer Assessment 1
It is now possible to collect a large amount of data about personal movement
using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone
Up. These type of devices are part of the "quantified self" movement - a group
of enthusiasts who take measurements about themselves regularly to improve their
health, to find patterns in their behavior, or because they are tech geeks. But
these data remain under-utilized both because the raw data are hard to obtain
and there is a lack of statistical methods and software for processing and
interpreting the data.

This assignment makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day. The data
consists of two months of data from an anonymous individual collected during
the months of October and November, 2012 and include the number of steps taken
in 5 minute intervals each day.  

The following code assumes you have downloaded and unzipped the file
activity.csv into your working directory.

## Loading and preprocessing the data
For this assignment we will be using the dplyr, ggplot2, and lattice packages
in addition to base R, so the first activity is to load the packages into
memory.
```{r Load Packages, results='hide', message=FALSE, warning=FALSE}
require(dplyr)
require(ggplot2)
require(lattice)
```
As a first glance at the data, we will load the raw data into a data.frame and
then copy it to another data.frame with rows containing NA removed so we can
perform some exploratory analysis
```{r Read Data}
data <- read.csv("activity.csv", head = TRUE, stringsAsFactors = FALSE)
clean.data <- na.omit(data)
```
Looking at the data shows it has three (3) rows, with number of steps, the
date, and the 5 minute interval that the row represents.  For the first part
of the assignment we need to look at the total steps per day.  We will use the
dplyr package to first group the data by the date, and then use the summarise()
function to sum up the number of steps based on the grouping.  We will then
plot the summations on a histogram in order to see how steps per day are
distributed.  We will plot the histogram using both the base R and ggplot2
(since peer assessment may be on comparison to ggplot2 settings.)
```{r Grouping and Summarizing, message = FALSE}
days <- group_by(clean.data, date)
total_days <- summarise(days, total = sum(steps, na.rm = TRUE))
hist(total_days$total, col = "green", xlab = "Total Steps Taken per Day",
     main = "Histogram of Total Steps in a Day", ylim = c(0,20), breaks = 20)
ggplot(total_days, aes(x = total)) + geom_histogram() + 
     labs(x = "Total Steps Taken per Day", y = "Frequency",
          title = "Histogram of Total Steps in a Day")
```
  
The histograms show that the Total Number of Steps appear to be Normally
(Gaussian) distributed.  With only sixty (60) data points, it is reasonable to
make that observation. More data would smooth out the curve.

## What is mean total number of steps taken per day?
Using the mean and median functions, we can calculate the centrality
characteristics of the Total Number of Steps.  We should also look at the
Standard Deviation to get a sense of how the data is distributed around the
mean.
```{r Data Descriptives}
paste("Mean =",mean(total_days$total), ";", "Median =", median(total_days$total),
      ";", "Standard Deviation =", sd(total_days$total), sep=" ")
```
The mean and median are practically the same, reinforcing the notion that the
Total Number of Steps is Normally Distributed.

## What is the average daily activity pattern?
Next we need to look at the daily pattern.  Again using dplyr, we group the
data by interval and get an average number of steps over the entire two months.
We then take the averages and plot them to get a profile of average steps over
a 'day'. 
```{r Daily Activity}
interval <- group_by(clean.data, interval)
interval_avg <- summarise(interval, avg = mean(steps))
plot(interval_avg$interval,interval_avg$avg, type = "l", xlab = "Interval (5 min increments)",
     ylab = "Average Number of Steps", main = "Average Number of Steps per Interval over Time")
```

The plot shows little to no activity from 2100 - 0500, with a peak near 200
around 0830, and sustained activity for the rest of the time. Using the grep()
and max() functions, we can find the maximum point and the points on either
side of it.
```{r Max Steps}
max_point <- grep((max(interval_avg$avg)), interval_avg$avg)
interval_avg[(max_point-1):(max_point+1),]
```

## Imputing missing values
Returning to the 'raw' data set we note that there are several NAs.  We use the
summary() function to look at the entire data.frame and note that all NAs occur
only in the steps column.  The function sum(is.na()) can also be used to just
get NAs in the data.frame, but gives no additional information.
```{r Data Summary}
summary(data)
sum(is.na(data))
```
  
To fill in the missing data, we will take the data.table that has the average
steps taken per interval and use it to fill in the missing data, hence days
that have all missing steps will be imputed with an 'average' days worth of
steps.  The following code creates a logical vector, missing_data, based on if
the data is missing and there is an NA.  We will create a new data.frame,
filled.data, that will have the missing data filled in while preserving the
state of the original data.frame in data.  The for loop goes thorugh each row
of the data.frame and if the logical element is TRUE, reads the interval, and
then pulls the average value from the interval_avg data.table and replaces the
NA in the filled data.frame with the value. 
```{r Imputing NAs, warning = FALSE}
missing_data <- is.na(data$steps)
filled.data <- data
for (i in 1:nrow(filled.data)) {
  if(missing_data[i] == TRUE) {
    missing_interval = filled.data[i,3]
    avg_index <- grep(missing_interval, interval_avg[,1])
    filled.data$steps[i] = interval_avg[avg_index,2]
  }
}
```

Repeating the same processing the on the filled data.frame as we did with the
cleaned one, we group the steps together for each day and plot the histogram. 
The original histogram is repeated for comparison.  We then calculate the Mean,
Median, and Standard Deviation for the filled data.frame.

```{r Filled Comparison, fig.height = 10}
filled.days <- group_by(filled.data, date)
filled.total_days <- summarise(filled.days, total = sum(steps))
par(mfrow = c(2,1))
hist(total_days$total, col = "green", xlab = "Total Steps Taken per Day",
     main = "Histogram of Total Steps in a Day", ylim = c(0,20), breaks = 20)

hist(filled.total_days$total, col = "green", xlab = "Total Steps Taken per Day",
     main = "Histogram of Total Steps in a Day (Filled Data)", ylim = c(0,20),
     breaks = 20)

paste("Mean =", mean(filled.total_days$total), ";", "Median =", 
      median(filled.total_days$total),";", "Standard Deviation =", 
      sd(filled.total_days$total), sep =" ")
```

The histograms show that our imputing only changed the bin that contained the
Mean days, which makes sense if we made NA days into average days.  This did
nothing to the Mean, and brought the Median slightly closer to the Mean.  It
did reduce the Standard Deviation by almost 200 steps.

## Are there differences in activity patterns between weekdays and weekends?
The final factor to look at is how Weekdays compare to Weekends.  In order to
look at this feature we use the weekdays() and as.Date() functions to convert a
date into the appropriate category.  When then use the gsub() function to
convert Saturday and Sunday into Weekend and the rest into Weekday.  We then
subset the data.frame into two separate data.frames based on Weekday or
Weekend.  The functions group_by() and summarise() from dplyr are used as
before to get the total steps in each interval, but now we have them separated
by Weekday or Weekend.  We then add the factor for day to the data.tables and 
then rbind them.  Finally we use the lattice package to plot a scatterplot on
two (2) panels based on the day.

```{r Day Type Segregation and comparision}
filled.data$day <- weekdays(as.Date(filled.data[,2]))
filled.data$day <- gsub("Sunday", "weekend", filled.data$day)
filled.data$day <- gsub("Saturday", "weekend", filled.data$day)
filled.data$day <- gsub("Monday", "weekday", filled.data$day)
filled.data$day <- gsub("Tuesday", "weekday", filled.data$day)
filled.data$day <- gsub("Wednesday", "weekday", filled.data$day)
filled.data$day <- gsub("Thursday", "weekday", filled.data$day)
filled.data$day <- gsub("Friday", "weekday", filled.data$day)

filled.data.weekday <- subset(filled.data, day == "weekday")
filled.data.weekend <- subset(filled.data, day == "weekend")

interval.weekday <- group_by(filled.data.weekday, interval)
interval.weekend <- group_by(filled.data.weekend, interval)
interval.weekday_avg <- summarise(interval.weekday, avg = mean(steps))
interval.weekday_avg$day <- "weekday"
interval.weekend_avg <- summarise(interval.weekend, avg = mean(steps))
interval.weekend_avg$day <- "weekend"

interval.week_avg <- rbind(interval.weekday_avg, interval.weekend_avg)
xyplot(interval.week_avg$avg ~ interval.week_avg$interval | interval.week_avg$day,
       layout = c(1,2), type = "l", xlab = "Interval (5 min increments)",
       ylab = "Average Number of Steps",
       main = "Average Number of Steps per Interval over Time")
```
  
Both plots show a peak of steps at the same time, but for Weekdays there is a
sudden increase in steps from zero with level activity until the peak.  The
Weekend data shows a more gradual buildup to a smaller magnitude peak.  Both
factors have sustained steps after the peak, but the weekend has a higher
average of steps than the weekday.  Both plots show steps decreasing gradually
at the same time and at the same rate.